---
title: "japan_analysis"
output: html_document
---
## Analysis in R

his R-markdown document contains the analysis part of the Japanese datasets

### Version of R and R-studio used in this R markdown. 

**R** = version 3.6.1
**R studio** = version 1.2.1335

### setting the working directory

```{r}
setwd("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\Japan")
list.files()
```

### loading the packages

```{r}

# install.packages("phyloseq")
# install.packages("vegan")s
#install.packages("DESeq2")
# install.packages("ggplot2")
# install.packages("dendextend")
# install.packages("tidyr")
# install.packages("viridis")
# install.packages("reshape")
# install.packages("iNEXT")
# install.packages("rgr")

library("phyloseq")
library("vegan")
library("DESeq2")
library("ggplot2")
#library("dendextend")
library("tidyr")
library("viridis")
library("reshape")
library("dplyr")
library("iNEXT")
library("rgr")
library("dendextend")
```


## Checking the package version 

```{r}
packageVersion("phyloseq")
packageVersion("vegan")
packageVersion("DESeq2")
packageVersion("ggplot2") 
#packageVersion("dendextend")
packageVersion("tidyr")
packageVersion("viridis")
packageVersion("reshape")
packageVersion("iNEXT")
packageVersion("rgr")
packageVersion("dendextend")
```


## Reading the exported data from the directory

```{r}

#Load ASV table
count_tab_japan <- read.table("ASVs_japan.tsv", header=T, row.names=1, check.names=F, sep="\t")
rownames(count_tab_japan) <- gsub(" ", "", rownames(count_tab_japan)) # Changin the format of rownames in ASV table from "ASV XXX _" to "ASVXXX_"

#Load taxonomy table
tax_tab_japan <- as.matrix(read.table("ASVs_japan_Taxa.tsv", header=T, row.names=1, check.names=F, sep = "\t"))

#Load sample matadata file
sample_info_tab_japan <- read.csv("SraRunTable_japan_without100m.txt", header = T, row.names = NULL, check.names = F, sep = "\t")
row.names(sample_info_tab_japan) <- sample_info_tab_japan$run_name # Assign rownames from one of the column "run_name"
sample_info_tab_japan <- sample_info_tab_japan[, -1] # Remove the column since it is assigned to the rownames

#Check content of the metadata file
sample_info_tab_japan
```

- **sep** = a string used to separate columns. using sep = "\t" gives tab deliminated output
- **quote** = a logical value or a numeric vector. If TRUE, any character or factor coluns will be surrounded by double quotes. If FALSE, nothing is quoted/
- **col.names** = either a logical value indicating whether the column names of x are to be written along with x, or a character vector of column names to be written. 
- **row.names** = a vector of row names. Setting it to 1 means the name of samples/ASV number is listed in the first column of the table
- **header** = TRUE if a table contains header for columns. 

## Assigning contigency tables and sample metadata into phyloseq object
```{r}

#read ASV table into OTU_table class object
otu_1_japan <- as.matrix(count_tab_japan)
otu_jp <- t(otu_1_japan)
otu_2_japan <- decostand(otu_1_japan, MARGIN = 2, method = "hellinger" )# transpose table for hellinger transformation
otu_i_japan <- otu_table(otu_1_japan, taxa_are_rows = T)
otu_d_japan <- otu_table(otu_2_japan, taxa_are_rows = T)
rm(otu_1_japan)
rm(otu_2_japan)

#read taxonomy table into tax_table class object
TAX_japan <- tax_table(tax_tab_japan)

#read metadata file into sample_data class object
samples_japan <- sample_data(sample_info_tab_japan)

#Construct phyloseq object from the component objects above. 
Ps_japan <- phyloseq(otu_i_japan, TAX_japan, samples_japan)
Ps_hellinger_japan <- phyloseq(otu_d_japan, TAX_japan, samples_japan)

#Check content of the object
Ps_japan
Ps_hellinger_japan

#Change column names to sample names
PS_japan_sample_name <- merge_samples(Ps_japan, "sample_name") # This is only neccesary if you are not using plotting functions by phyloseq (e.g. dendrogram), otherwise, the name of samples can be manipulated through a parameter in phyloseq plotting function

```


-- **merge_samples()** = merge/agglomerate the sample indices of a phyloseq object according to a categorical variable contained in a sample_data or a provided factor. group = a single character string mathcin a variable name in the correspoding sample date of x.

## Plotting Chao1 richness estimator and Shannon diversity estimator. 

Chao1 estimator is measured by the number of observed species plus number of singletons divided by two times of number of doubletons. Chao1 estimates the number of species that are represented only by a single indivudal or by two inviduals in that sample based on abundances.

Shannon diversity estimator shows both richness and evenness of species abundance. In other words, if a sample has high shannon diversity index, the sample is thought to be more 'diverse' compared to those samples with lower diversity index 

Simpson's diversity index is a measure of diversity which takes into account the number of species present, as well as the relative abundance of each species. 

```{r}
#plot alpha diversity by samples

#Defining date format 
sample_data(Ps_japan)$sample_name <- factor(sample_data(Ps_japan)$sample_name, levels = sample_data(Ps_japan)$sample_name[order(sample_data(Ps_japan)$Collection_Date)])

# PLot
plot_richness(Ps_japan, measures = c("Chao1", "Shannon", "Simpson"), x = "sample_name", color = "Collection_Date") + geom_point(size = 4) + theme(axis.text.x = element_text(size = 8))
```


```{r}
# plot alpha diversity by Collection date

plot_richness(Ps_japan, measures = c("Chao1", "Shannon", "Simpson"), x="Collection_Date", color = "Depth") + labs(x = "Collection Date") + geom_point(size = 4) + 
   scale_color_discrete(breaks=c("0m", "5m", "10m", "50m", "200m")) #Reordering label items
```

## iNEXT rarefaction curve

iNEXT function only allws data frame or contigency table as input, so if samples were grouped or names of columns were changed in phyloseq, it is important to transform a phyloseq oject into a data frame. For my case, column names of the original table are accession numbers from NCBI and the names need to be changed to individual sediment depth of sample extraction with name of sampling site. 
```{r}
# Exporting a count table with desired sample names from a phyloseq object 
otu_table_site_japan <- as(otu_table(PS_japan_sample_name), "matrix")
if(!taxa_are_rows(PS_japan_sample_name)){otu_table_site_japan <- t(otu_table_site_japan)}
otu_table_site_df_japan <- as.data.frame(otu_table_site_japan)

# Check the content
otu_table_site_df_japan

```

```{r}
# rarefaction curve using iNEXT 
out_japan <- iNEXT(otu_table_site_df_japan, q=0, datatype = "abundance")
out_japan$DataInfo
ggiNEXT(out_japan, type=1, se= TRUE, facet.var = "order", color.var = "site") + theme(legend.text = element_text(size = 7)) + ylab("Number of ASVs")
ggiNEXT(out_japan, type=2, se = TRUE, facet.var = "order", color.var = "site") + theme(legend.text = element_text(size = 7)) + ggtitle("Sample completeness curve")
ggiNEXT(out_japan, type=3, se = TRUE, facet.var = "order", color.var = "site") + theme(legend.text = element_text(size = 7)) + ggtitle("Coverage based curve")
```

There are three types of curves plotted above.

- **Type 1** = Sample- size based R/E curve. This curve plots diversity estimates with confidence intervals as a function of sample size up to double the reference sample size.
- **Type 2** = sample completeness curve with confidence intervals. This curve plots the sample coverage with respect to sample size for the same range with type 1 plot
- **Type 3** = Coverage based R/E curve, This curve plots the diversity estimates with confidence intervals as a function of sample coverage up to the maximum coverage obrained from the maximum size described in type 1
- **knots** = an integer specifying the numver of equally-spaced knots between size 1 andthe endpoint. Since my data set is large and the compational time takes forever, the knots parameter is set to 100 to draw the rarefaction curves. 
- **conf** = a positive number <1 specifying the level of condence interval. I do not wish to calculate the confidence interval. Therefore, the value is set to NULL. 
- **se** = if TRUE, it plots confidence intervals. 
- **theme()** = Changing legend text font size. 
- **ggtitle()** = add title to plot. 
-

## Community Compositions

Ratio transformations capture the relationships between each ASVs in the dataset and these ratios are the same whether the data are counts or proportions. Taking the logarithm of these ratios makes the data symmetric and linearly related and places the data in a log ratio coordinate space. 

```{r}

# export a count table with the right column name
merge_jp <- merge(count_tab_japan, tax_tab_japan, by = 0, sort = FALSE)

# Aggregate ASV counts for Phylum
clr_phylum_jp <-  merge_jp %>% 
  dplyr::select(2:17, 20) %>% # select() extracts columns from input number. 
  group_by(Phylum) %>%  # aggreagate counts by phylum
  summarise_all(funs(sum)) 

# Removing first row since the taxonomy name is unknown.
clr_phylum_jp <- clr_phylum_jp[-1,] 

# Make taxonomy name as rownames
row.names(clr_phylum_jp) <- clr_phylum_jp$Phylum
clr_phylum_jp$Phylum <- NULL

# make samples into rows

clr_phylum_jp_t <- t(clr_phylum_jp)

# if the value in the table is 0, add a small constant 0.1 to make sure we can take the log
clr_phylum_jp_t[clr_phylum_jp_t == 0] <- 0.1

# CLR transformation on OTU table
# CLR = taking log from ratio calculated by proportion of each ASVs within a sample

rm(clr_phylum_jp)
clr_phylum_jp <- t(clr(clr_phylum_jp_t, ifwarn = FALSE))
rm(clr_phylum_jp_t)

# Repeat same steps for Class, Order, and Family. In this block, I will use source() to reduce redundancy

source("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\Japan\\clr_class_order_family.R")
```

- **by** = specifications of the columns used for merging. By default the data frames are merged on the columns with names they both have. In this case, rownames from the two tables need to match, so **0** is used in this parameter.
- **sort** = logical. Should the result be sorted on the by columns? 
- **ifwarn** = by default ifwarn = TRUE which generates a reminder/warning that when carrying out a centred log-ratio transformation all the data/parts must be in the same measurement units. The message can be suppressed by setting ifwarn = FALSE.

Now we can plot
```{r}
# plot

source("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\Japan\\plot_community_compositions.R")

```

## Hiracheal clustering dendrogram
```{r}

colDate <- get_variable(Ps_hellinger_japan, "Collection_Date")
Ps_sample_name_jp <- merge_samples(Ps_hellinger_japan, "sample_name")

#Caluclate distances
d_jp <- phyloseq::distance(Ps_sample_name_jp, method = "bray", type = "samples")

# Hierarchical cluster analysis on a set of dissimilarities
p.hclust_jp <- as.dendrogram(hclust(d_jp, method = "average"))

# color scheme
color_jp <- c("blue", "orange", "red", "green", "blue", "blue", "blue", "red", "red", "red", "green", "green", "green", "orange", "orange", "orange")
labels_colors(p.hclust_jp) <- color_jp 

# plot
plot(p.hclust_jp, hang = -1, cex.lab = 4, cex.axis = 0.8)


```

## NMDS ordination plot


NMDS is a way to condese information from mulidementional data into a 2D representation or ordination. The closer two points are, the more similar the corresponding samples are with respect to the variables that went into making the NMDS plot. 

```{r}

# Hellinger transformation
NMDS_japan <- decostand(otu_jp, MARGIN = 2, method = "hellinger" )

#Transpose the table.
example_NMDS <- vegan::metaMDS(NMDS_japan, k=2, distance = "bray")

# stress plot
stressplot(example_NMDS)
```

- **MARGIN** = Margin, if default is not accept. "1" = rows, and "2" = columns of x
- **k** = Number of dimensions. The number od points n should be n>2*k+1 and prefeably higher in non-metric MDS
- **distance** = Dissimilarity index used in vegdist. We will use bray curtis dissimilarity index. 

OKay now moving on to plotting the points


```{r}

# Overview of our data
plot(example_NMDS)

#extract NMDS scores (x and y coordinates)
data.scores = as.data.frame(scores(example_NMDS))

#add columns to data frame
data.scores$Collection_Date = sample_info_tab_japan$Collection_Date
data.scores$Depth = sample_info_tab_japan$Depth
data.scores$sample_name = sample_info_tab_japan$sample_name

# Check the content of the metadata
head(data.scores)

# plotting the NMDS.
NMDS_new_jp <- ggplot(data.scores, aes(x = NMDS1, y= NMDS2)) + 
  geom_point(size = 9, aes( shape = Depth, color = Collection_Date)) + # Choosing mapping for the plot
  scale_shape_manual(values = c(16, 15, 17, 3, 8)) + # Specifying shapes for each depth
  scale_shape_discrete(breaks=c("0m", "5m", "10m", "50m", "200m")) # Reordeing Legend items
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
    legend.text = element_text(size = 12, face ="bold", colour ="black"), 
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank()) + 
    labs(x = "NMDS1", color = "Collection Date", y = "NMDS2", shape = "Depth (m)")  # Label formatting
 

NMDS_new_jp
```


## Redundancy analysis 

Redundancy analysis (RDA) is a method to extract and summarise the variation in a set of response variables that can be explained by a set of explanatory 
variables. More accurately, RDA is a direct gradient analysis technique which summarises linear relationships between components of response variables that are 
"redundant" with (i.e. "explained" by) a set of explanatory variables.

```{r}

# Reorder sample name by date
sample_data(Ps_hellinger_japan)$sample_name <- factor(sample_data(Ps_hellinger_japan)$sample_name, levels = sample_data(Ps_hellinger_japan)$sample_name[order(sample_data(Ps_hellinger_japan)$Collection_Date)])


# ordinate
rda_japan <- ordinate(Ps_hellinger_japan, method = "RDA", distance = "bray", formula = ~ temperature + Salinity + Oxygen + Nitrate + CHL_a)

# Plot

RDA_japan <- plot_ordination(Ps_hellinger_japan, rda_japan, type = "sample", color= "sample_name", shape = "Collection_Date" )
RDA_japan



# Now add the environmental variables as arrows
arrowmat_japan = vegan::scores(rda_japan, display = "bp")

# Add labels, make a data.frame
arrowdf_japan <- data.frame(labels = rownames(arrowmat_japan), arrowmat_japan)

# Define the arrow aesthetic mapping
arrow_map = aes(xend = RDA1, yend = RDA2, x = 0, y = 0, shape = NULL, color = NULL, label = labels)
arrowhead = arrow(length = unit(0.05, "npc"))
label_map <- aes(x = 0.85 * RDA1, 
    y = 1.2 * RDA2, 
    shape = NULL, 
    color = NULL, 
    label = labels)

# Make a new graphic
RDA_japan + 
  geom_point(size = 5) + 
  geom_segment(
    mapping = arrow_map, 
    size = .5, 
    data = arrowdf_japan, 
    color = "gray", 
    arrow = arrowhead
  ) + 
  geom_text(
    mapping = label_map, 
    size = 5,  
    data = arrowdf_japan, 
    show.legend = FALSE
  ) + 
    theme(legend.text = element_text(size = 10, face ="bold", colour ="black")) +
    guides(fill=guide_legend(ncol=2))

```
