---
title: "EEC_analysis"
output: html_document
---


## Analysis in R

This R-markdown document contains the analysis part of the EEC datasets

### Version of R and R-studio used in this R markdown. 

**R** = version 3.6.1
**R studio** = version 1.2.1335

### setting the working directory

```{r}
setwd("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\EEC")
list.files()
```

### loading the packages

```{r}

# install.packages("phyloseq")
# install.packages("vegan")s
#install.packages("DESeq2")
# install.packages("ggplot2")
# install.packages("dendextend")
# install.packages("tidyr")
# install.packages("viridis")
# install.packages("reshape")
# install.packages("iNEXT")
# install.packages("dendextend")
# install.packages("rgr")

library("phyloseq")
library("vegan")
library("DESeq2")
library("ggplot2")
#library("dendextend")
library("tidyr")
library("viridis")
library("reshape")
library("dplyr")
library("iNEXT")
library("rgr")
library("dendextend")
```


## Checking the package version 

```{r}
packageVersion("phyloseq")
packageVersion("vegan")
packageVersion("DESeq2")
packageVersion("ggplot2") 
#packageVersion("dendextend")
packageVersion("tidyr")
packageVersion("viridis")
packageVersion("reshape")
packageVersion("iNEXT")
packageVersion("rgr")
packageVersion("dendextend")
```


## Reading the exported data from the directory
I create two phyloseq object for 2011-2013 set and 2013-2015 set and merge them together.
```{r}
#=====================================================================
# 2011- 2013
#====================================================================
# Load ASV table

count_tab_eec_2011_2013 <- read.table("ASVs_EEC_2011_2013.tsv", header=T, row.names=1, check.names=F, sep="\t")
count_tab_eec_2011_2013 <- count_tab_eec_2011_2013[, c(-5, -14)] # Removing two samples

# Changin the format of rownames in ASV table from "ASV XXX _" to "ASVXXX_"
rownames(count_tab_eec_2011_2013) <- gsub(" ", "", rownames(count_tab_eec_2011_2013)) # Chage the format of rownames from "ASV 1 _" to "ASV1_". 

#Load taxonomy table
tax_tab_eec_2011_2013 <- as.matrix(read.table("ASVs_eec_2011_2013_Taxa.tsv", header=T, row.names=1, check.names=F, sep = "\t"))

#Load sample matadata file
sample_info_tab_2011 <- read.csv("EEC_metadata_2011.txt", header = T, row.names = NULL, check.names = F, sep = "\t")
row.names(sample_info_tab_2011) <- sample_info_tab_2011$Run # Assign rownames from one of the column "Run"
sample_info_tab_2011 <- sample_info_tab_2011[,-1] # Remove the column since it is assigned to the rownames

#Check content of the metadata file
sample_info_tab_2011

#====================================================================
# 2013 - 2015
#====================================================================
#Load ASV table
count_tab_eec_2013_2015 <- read.table("ASVs_EEC_2013_2015.tsv", header=T, row.names=1, check.names=F, sep="\t")

# Changin the format of rownames in ASV table from "ASV XXX _" to "ASVXXX_"
rownames(count_tab_eec_2013_2015) <- gsub(" ", "", rownames(count_tab_eec_2013_2015))

#Load taxonomy table
tax_tab_eec_2013_2015 <- as.matrix(read.table("ASVs_eec_2013_2015_Taxa.tsv", header=T, row.names=1, check.names=F, sep = "\t"))

#Load sample matadata file
sample_info_tab_2013 <- read.csv("EEC_metadata_2013.txt", header = T, row.names = NULL, check.names = F, sep = "\t")
row.names(sample_info_tab_2013) <- sample_info_tab_2013$Run
sample_info_tab_2013 <- sample_info_tab_2013[,-1]

#Check content of the metadata file
sample_info_tab_2013

```

- **sep** = a string used to separate columns. using sep = "\t" gives tab deliminated output
- **quote** = a logical value or a numeric vector. If TRUE, any character or factor coluns will be surrounded by double quotes. If FALSE, nothing is quoted/
- **col.names** = either a logical value indicating whether the column names of x are to be written along with x, or a character vector of column names to be written. 
- **row.names** = a vector of row names. Setting it to 1 means the name of samples/ASV number is listed in the first column of the table
- **header** = TRUE if a table contains header for columns. 


## Assigning contigency tables and sample metadata into phyloseq object
```{r}


#read ASV table into OTU_table class object
otu_1_2011 <- as.matrix(count_tab_eec_2011_2013)
otu_1_2013 <- as.matrix(count_tab_eec_2013_2015)

# transpose table for hellinger transformation
otu_2011 <- t(otu_1_2011)
otu_2_2011 <- decostand(otu_1_2011, MARGIN = 2, method = "hellinger" )
otu_i_2011 <- otu_table(otu_1_2011, taxa_are_rows = T)
otu_d_2011 <- otu_table(otu_2_2011, taxa_are_rows = T)
rm(otu_1_2011)
rm(otu_2_2011)

otu_2013 <- t(otu_1_2013)
otu_2_2013 <- decostand(otu_1_2013, MARGIN = 2, method = "hellinger" )
otu_i_2013 <- otu_table(otu_1_2013, taxa_are_rows = T)
otu_d_2013 <- otu_table(otu_2_2013, taxa_are_rows = T)
rm(otu_1_2013)
rm(otu_2_2013)

#read taxonomy table into tax_table class object
TAX_2011 <- tax_table(tax_tab_eec_2011_2013)
TAX_2013 <- tax_table(tax_tab_eec_2013_2015)

#read metadata file into sample_data class object
samples_2011 <- sample_data(sample_info_tab_2011)
samples_2013 <- sample_data(sample_info_tab_2013)

#Construct phyloseq object from the component objects above. 
Ps_2011 <- phyloseq(otu_i_2011, TAX_2011, samples_2011)
Ps_hellinger_2011 <- phyloseq(otu_d_2011, TAX_2011, samples_2011)

Ps_2013 <- phyloseq(otu_i_2013, TAX_2013, samples_2013)
Ps_hellinger_2013 <- phyloseq(otu_d_2013, TAX_2013, samples_2013)

#Check content of the object
Ps_2011
Ps_hellinger_2011
Ps_2013
Ps_hellinger_2013

```

## Merging two phyloseq objects

```{r}
# Merge two phyloseq objects
Ps_combined <- merge_phyloseq(Ps_2011, Ps_2013)
Ps_EEC <- tax_glom(Ps_combined, "Genus")
rm(Ps_combined)

# Same for hellinger transformed objects
Ps_combined <- merge_phyloseq(Ps_hellinger_2011, Ps_hellinger_2013)
Ps_EEC_hellinger <- tax_glom(Ps_combined, "Genus")
rm(Ps_combined)
```

- **merge_phyloseq()** = Takes a comma-separated list of phyloseq objects as arguments, and returns the most-comprehensive single phyloseq object possible.


## Plotting Chao1 richness estimator and Shannon diversity estimator. 

Chao1 estimator is measured by the number of observed species plus number of singletons divided by two times of number of doubletons. Chao1 estimates the number of species that are represented only by a single indivudal or by two inviduals in that sample based on abundances.

Shannon diversity estimator shows both richness and evenness of species abundance. In other words, if a sample has high shannon diversity index, the sample is thought to be more 'diverse' compared to those samples with lower diversity index 

Simpson's diversity index is a measure of diversity which takes into account the number of species present, as well as the relative abundance of each species. 

```{r}
#plot alpha diversity by samples

test <-Ps_EEC

#Defining date format 

date_EEC <- as.Date(sample_data(Ps_EEC)$collection_date, format = '%m/%d/%Y')
sample_data(test)$collection_date  <- as.factor(date_EEC)
sample_data(test)$year <- format(date_EEC, "%Y") # This is to add points by year
                                               
# Plot

plot_richness(test, measures = c("Chao1", "Shannon", "Simpson"), x = "collection_date", color = "year") + geom_point(size = 4) + theme(axis.text.x = element_text(size = 8)) +
    theme(legend.title = element_text(size = 14),
        legend.text = element_text(size = 12, face = "bold"),
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 0.6)
  ) +
  labs(color = "Collection Year", x = "Collection Date")
```


```{r}
#plot alpha diversity by samples colored by season
plot_richness(test, measures = c("Chao1", "Shannon", "Simpson"), x = "collection_date", color = "season") + geom_point(size = 4) + theme(axis.text.x = element_text(size = 8)) +
 theme(legend.title = element_text(size = 14),
        legend.text = element_text(size = 12, face = "bold"),
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 0.6)
  ) +
  labs(color = "Season", x = "Collection Date")
```




## iNEXT rarefaction curve

iNEXT function only allws data frame or contigency table as input, so if samples were grouped or names of columns were changed in phyloseq, it is important to transform a phyloseq oject into a data frame. For my case, column names of the original table are accession numbers from NCBI and the names need to be changed to individual sediment depth of sample extraction with name of sampling site. 
```{r}

#Change name of column by sample names
sample_data(Ps_EEC)$collection_date <- as.Date(sample_data(Ps_EEC)$collection_date, format = '%m/%d/%Y')
PS_EEC_sample_name <- merge_samples(Ps_EEC, "collection_date")

otu_table_site_EEC <- as(otu_table(PS_EEC_sample_name), "matrix")
if(!taxa_are_rows(PS_EEC_sample_name)){otu_table_site_EEC <- t(otu_table_site_EEC)}
otu_table_site_df_EEC <- as.data.frame(otu_table_site_EEC)

# Check the content
otu_table_site_df_EEC

```

```{r}
# rarefaction curve using iNEXT
out_EEC <- iNEXT(otu_table_site_df_EEC, q=0, datatype = "abundance")
out_EEC$DataInfo
ggiNEXT(out_EEC, type=1, se= TRUE, facet.var = "order", color.var = "site") + theme(legend.text = element_text(size = 7)) + ylab("Number of ASVs") +
 theme(legend.title = element_text(size = 14),
        legend.text = element_text(size = 12, face = "bold"),
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 0.6)
  ) +
  guides(color=guide_legend(title="Guides", ncol=length(levels(factor(out_EEC$DataInfo$site)))))


ggiNEXT(out_EEC, type=2, se = TRUE, facet.var = "order", color.var = "site") + theme(legend.text = element_text(size = 7)) + ggtitle("Sample completeness curve")
ggiNEXT(out_EEC, type=3, se = TRUE, facet.var = "order", color.var = "site") + theme(legend.text = element_text(size = 7)) + ggtitle("Coverage based curve")
```


- **Type 1** = Sample- size based R/E curve. This curve plots diversity estimates with confidence intervals as a function of sample size up to double the reference sample size.
- **Type 2** = sample completeness curve with confidence intervals. This curve plots the sample coverage with respect to sample size for the same range with type 1 plot
- **Type 3** = Coverage based R/E curve, This curve plots the diversity estimates with confidence intervals as a function of sample coverage up to the maximum coverage obrained from the maximum size described in type 1
- **knots** = an integer specifying the numver of equally-spaced knots between size 1 andthe endpoint. Since my data set is large and the compational time takes forever, the knots parameter is set to 100 to draw the rarefaction curves. 
- **conf** = a positive number <1 specifying the level of condence interval. I do not wish to calculate the confidence interval. Therefore, the value is set to NULL. 
- **se** = if TRUE, it plots confidence intervals. 
- **theme()** = Changing legend text font size. 
- **ggtitle()** = add title to plot. 
-

## Community Compositions

Ratio transformations capture the relationships between each ASVs in the dataset and these ratios are the same whether the data are counts or proportions. Taking the logarithm of these ratios makes the data symmetric and linearly related and places the data in a log ratio coordinate space. 

```{r}
# export the count table with right column names
count_EEC_correct_names <- as(otu_table(PS_EEC_sample_name), "matrix")  # extract a count table from the phyloseq object

# export the taxonomy table from the phyloseq object
tax_EEC <- as(tax_table(PS_EEC_sample_name), "matrix")

# Merge count table and taxa table
merge_EEC <- merge(t(count_EEC_correct_names), tax_EEC, by = 0, sort = FALSE)

# Aggregate ASV counts for Phylum
clr_phylum_EEC <-  merge_EEC %>% 
  dplyr::select(2:29, 32) %>%  # select() extracts columns from input number.  
  group_by(Phylum) %>% # aggreagate counts by phylum
  summarise_all(funs(sum)) 

# Checking sum of the counts before and after the aggregation. 

colSums(clr_phylum_EEC[,-1]) # sum of aggregated table
colSums(t(count_EEC_correct_names)) # sum of original table

# Make taxonomy name as rownames
row.names(clr_phylum_EEC) <- clr_phylum_EEC$Phylum
clr_phylum_EEC$Phylum <- NULL

# make samples into rows

clr_phylum_EEC_t <- t(clr_phylum_EEC)

# if the value in the table is 0, add a small constant 0.1 to make sure we can take the log
clr_phylum_EEC_t[clr_phylum_EEC_t == 0] <- 0.1

# CLR transformation on OTU table
# CLR = taking log from ratio calculated by proportion of each ASVs within a sample

rm(clr_phylum_EEC)
clr_phylum_EEC <- t(clr(clr_phylum_EEC_t, ifwarn = FALSE))
rm(clr_phylum_EEC_t)

# Repeat same steps for Class, Order, and Family. In this block, I will use source() to reduce redundancy

source("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\EEC\\clr_class_order_family.R")
```
- **by** = specifications of the columns used for merging. By default the data frames are merged on the columns with names they both have. In this case, rownames from the two tables need to match, so **0** is used in this parameter.
- **sort** = logical. Should the result be sorted on the by columns? 
- **ifwarn** = by default ifwarn = TRUE which generates a reminder/warning that when carrying out a centred log-ratio transformation all the data/parts must be in the same measurement units. The message can be suppressed by setting ifwarn = FALSE.


**Cross checking**
When plotting comunity compositions by each taxa rank (phylum, class, order, and family), the counts that corespond to identical taxonomy must be aggregated in order to avoid redundancy in the dataset. The original data table is filtered into each taxa rank with aggregated ASV counts and the data is stored in a object. Correct taxonomy names and its number of counts are manually cross checked by looking at first few rows in the object and the original data table. 

Now we can plot. 

```{r}
# plot

source("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\EEC\\plot_community_compositions.R")

```


## Hiracheal clustering dendrogram
```{r}

colDate <- get_variable(Ps_EEC_hellinger, "collection_date")
Ps_hellinger_EEC <- merge_samples(Ps_EEC_hellinger, "collection_date")

#Caluclate distances

d_EEC <- phyloseq::distance(Ps_hellinger_EEC, method = "bray", type = "samples")

# Hierarchical cluster analysis on a set of dissimilarities
p.hclust_EEC <- as.dendrogram(hclust(d_EEC, method = "average"))

# color scheme
color_EEC <- c("green", "green", "blue", "blue", "green", "green", "blue", "red", "red", "green", "blue", "blue", "red", "red", "red", "green", "green", "green", "red", "red", "red", "red", "green", "red", "red", "red", "red", "red")
labels_colors(p.hclust_EEC) <- color_EEC 

# plot
plot(p.hclust_EEC, hang = -1, cex.lab = 1, cex.axis = 0.8, ylab = "Bray-Curtis Dissimilarity")


```


## NMDS ordination plot

NMDS is a way to condese information from mulidementional data into a 2D representation or ordination. The closer two points are, the more similar the corresponding samples are with respect to the variables that went into making the NMDS plot. 

```{r}


NMDS_EEC <- decostand(count_EEC_correct_names, MARGIN = 2, method = "hellinger" )

#Transpose the table.
EEC_NMDS <- vegan::metaMDS(NMDS_EEC, k=2, distance = "bray")
rm(NMDS_EEC)

# stress plot
stressplot(EEC_NMDS)
```


- **MARGIN** = Margin, if default is not accept. "1" = rows, and "2" = columns of x
- **k** = Number of dimensions. The number od points n should be n>2*k+1 and prefeably higher in non-metric MDS
- **distance** = Dissimilarity index used in vegdist. We will use bray curtis dissimilarity index. 

OKay now moving on to plotting the points

```{r}
plot(EEC_NMDS)

#extract NMDS scores (x and y coordinates)
data.scores_EEC <- as.data.frame(scores(EEC_NMDS))

#add columns to data frame
data.scores_EEC$collection_date = sample_data(Ps_EEC)$collection_date
data.scores_EEC$year = format(date_EEC, "%Y") # This is to add points by year
data.scores_EEC$season = sample_data(Ps_EEC)$season

head(data.scores_EEC)

# plotting the NMDS.
NMDS_new_EEC <- ggplot(data.scores_EEC, aes(x = NMDS1, y= NMDS2)) + 
  geom_point(size = 7, aes( shape = as.factor(year), color = season )) + # Choosing mapping for the plot
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
    legend.text = element_text(size = 12, face ="bold", colour ="black"), 
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank()) + 
    labs(x = "NMDS1", color = "Season", y = "NMDS2", shape = "Collection Year") # Label formatting

NMDS_new_EEC
```



## Redundancy analysis 


Redundancy analysis (RDA) is a method to extract and summarise the variation in a set of response variables that can be explained by a set of explanatory 
variables. More accurately, RDA is a direct gradient analysis technique which summarises linear relationships between components of response variables that are 
"redundant" with (i.e. "explained" by) a set of explanatory variables.

```{r}

sample_data(Ps_EEC_hellinger)$collection_date <- as.factor(format(date_EEC, "%Y"))
# ordinate
rda_EEC <- ordinate(Ps_EEC_hellinger, method = "RDA", distance = "bray", formula = ~ Temperature + Salinity + Oxygen + POC + pH + PO4 + SiOH4 + PON + CHLa)

# Plot

RDA_EEC <- plot_ordination(Ps_EEC_hellinger, rda_EEC, type = "sample", color="season", shape = "collection_date")

RDA_EEC



# Now add the environmental variables as arrows
arrowmat_EEC = vegan::scores(rda_EEC, display = "bp")

# Add labels, make a data.frame
arrowdf_EEC <- data.frame(labels = rownames(arrowmat_EEC), arrowmat_EEC)

# Define the arrow aesthetic mapping
arrow_map = aes(xend = RDA1, yend = RDA2, x = 0, y = 0, shape = NULL, color = NULL, label = labels)
arrowhead = arrow(length = unit(0.05, "npc"))
label_map <- aes(x = 0.97 * RDA1, 
    y = 1.2 * RDA2, 
    shape = NULL, 
    color = NULL, 
    label = labels)

# Make a new graphic
RDA_EEC + 
  geom_point(size = 5) + 
  geom_segment(
    mapping = arrow_map, 
    size = .5, 
    data = arrowdf_EEC, 
    color = "gray", 
    arrow = arrowhead
  ) + 
  geom_text(
    mapping = label_map, 
    size = 5,  
    data = arrowdf_EEC, 
    show.legend = FALSE
  ) +
  geom_point(size = 4) + # Choosing mapping for the plot 
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), # axis text formatting
    legend.text = element_text(size = 12, face ="bold", colour ="black"), # legend text formatting
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), # legend position formatting
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), # axis title formatting
    legend.title = element_text(size = 12, colour = "black", face = "bold"), # legend title formatting
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), # change background color
    legend.key=element_blank()) + 
    labs(color = "Season", shape = "year" ) # Label formatting

```

## **MDS ordination plot**

```{r}

#Calculate distance
Ps_MDS_EEC <- ordinate(Ps_EEC_hellinger, "MDS", "bray")

#plot

#by dates
plot_ordination(Ps_EEC_hellinger, Ps_MDS_EEC, color = "season", shape = "collection_date") +
  geom_point(size = 4) + # Choosing mapping for the plot 
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), # axis text formatting
    legend.text = element_text(size = 12, face ="bold", colour ="black"), # legend text formatting
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), # legend position formatting
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), # axis title formatting
    legend.title = element_text(size = 12, colour = "black", face = "bold"), # legend title formatting
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), # change background color
    legend.key=element_blank()) + 
    labs(color = "Season", shape = "Collection year") # Label formatting



```
