---
title: "Peakpoc_metadata_average_import"
output: html_document
---
DADA2 processing of data for eastern English Channel study.

This R-markdown document includes a pipeline for data processing of amplicon data taken from the publication by [link](https://doi. org/10.1371/journal.pone.0196987).

### Version of all the programs and tools used n this R markdown. 

**R** = 3.6.1
**R studio** = 1.2.1335
**tidyr** = 1.0.0
**dplyr** = 1.4.2


### Loading required libraries

```{r}
library(tidyr)
library(dplyr)
```

### Importing data

```{r}
# Loading environmental data and metadata table

#Sea surface temperature
dsst <- read.csv("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\PeakPOC\\SST_IV.csv", sep =",", dec = ".", header=T, row.names = NULL)

# Ice concentration
dICEC <- read.delim("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\PeakPOC\\IV_20000101-20161231.txt", sep ="", dec = ".", header=T, row.names = NULL)

# CHlorophyll a
dCH1a_HGIV <- read.delim("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\PeakPOC\\IV_CMEMS-CHL-Arctic-oc5ci_20000101-20161231.txt", sep="", dec =".", header=T, row.names = NULL)

# Metadata 
metadata <- read.csv("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\PeakPOC\\Metadata_EUK_forMatomo.csv", sep=",", dec=".", header=T, row.names = NULL)

```

### Rearrange the data

```{r}
cols <- c("year", "month", "day")
cols2 <- c("YEAR", "MONTH", "DAY")
cols3 <- c("Year", "Month", "Day")

dsst$date <- apply( dsst[ , cols ] , 1 , paste , collapse = "-" )
dsst$date <- as.Date(dsst$date, format = "%Y-%m-%d")
dICEC$date <- apply( dICEC[ , cols2 ] , 1 , paste , collapse = "-" )
dICEC$date <- as.Date(dICEC$date, format = "%Y-%m-%d")
dCH1a_HGIV$date <- apply( dCH1a_HGIV[ , cols3 ] , 1 , paste , collapse = "-" )
dCH1a_HGIV$date <- as.Date(dCH1a_HGIV$date, format = "%Y-%m-%d")
```

### fix the concentrations values
```{r}
dsst$sst <- as.numeric(dsst$sst)
dICEC$ICEC <- as.numeric(dICEC$ICEC)
dCH1a_HGIV$CHL <- as.numeric(dCH1a_HGIV$CHL)
```

### get the mean according to the date

```{r}
#the function will store as a new column on the metadata table

for (beginning in metadata$collection_date_start){
  finish <- metadata %>% filter(collection_date_start == beginning) %>% select(collection_date_end)
  finish <- as.Date(finish$collection_date_end, "%Y-%m-%d")
  beginning <- as.Date(beginning, "%Y-%m-%d")

metadata$sst <- apply(metadata,1, function(z) {
  mean(dsst$sst[which(dsst$date >= z[beginning] & dsst$date <= z[finish])], na.rm=T)
}
)
metadata$dICEC <- apply(metadata, 1, function(z) {
  mean(dICEC$ICEC[which(dICEC$date >= z[beginning] & dICEC$date <= z[finish])], na.rm=T)
}
)

metadata$dCH1a_HGIV <- apply(metadata, 1, function(z) {
  mean(dCH1a_HGIV$CHL[which(dCH1a_HGIV$date >= z[beginning] & dCH1a_HGIV$date <= z[finish])], na.rm=T)
  
  
}
)
print(beginning)
  print(finish)
  print(mean(dsst$sst[which(dsst$date >= beginning & dsst$date <= finish)], na.rm=T))
  print(mean(dICEC$ICEC[which(dICEC$date >= beginning & dICEC$date <= finish)], na.rm=T))
  print(mean(dCH1a_HGIV$CHL[which(dCH1a_HGIV$date >= beginning & dCH1a_HGIV$date <= finish)], na.rm=T))
  rm(finish)
}


write.csv(metadata, file = "./metadata_EUK_new.csv")

```

