---
title: "peakpoc_analysis"
output: html_document
---

## Analysis in R

This R-markdown document contains the analysis part of the PeakPOC datasets

### Version of R and R-studio used in this R markdown. 

**R** = version 3.6.1
**R studio** = version 1.2.1335

### setting the working directory

```{r}
setwd("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\PeakPOC")
list.files()
```
 
### loading all the neccesary  packages

```{r}

# install.packages("phyloseq")
# install.packages("vegan")s
# install.packages("DESeq2")
# install.packages("ggplot2")
# install.packages("dendextend")
# install.packages("tidyr")
# install.packages("viridis")
# install.packages("reshape")
# install.packages("iNEXT")
# install.packages("rgr")
# install.packages("dendextend")


library("phyloseq")
library("vegan")
library("DESeq2")
library("ggplot2")
#library("dendextend")
library("tidyr")
library("viridis")
library("reshape")
library("dplyr")
library("iNEXT")
library("rgr")
library("dendextend")
library("data.table")
```


## Checking the package version 

```{r}
packageVersion("phyloseq")
packageVersion("vegan")
packageVersion("DESeq2")
packageVersion("ggplot2") 
#packageVersion("dendextend")
packageVersion("tidyr")
packageVersion("viridis")
packageVersion("reshape")
packageVersion("iNEXT")
packageVersion("rgr")
packageVersion("dendextend")
```


## Reading the exported data from the directory

```{r}

#Load ASV table
count_tab_peakpoc <- read.table("ASVs_peakpoc.tsv", header=T, row.names=1, check.names=F, sep="\t")
rownames(count_tab_peakpoc) <- gsub(" ", "", rownames(count_tab_peakpoc)) # Chage the format of rownames from "ASV 1 _" to "ASV1_". 
count_tab_peakpoc <- count_tab_peakpoc[,-20] # Removing the column for FEVI_3_Oben_3

#Load taxonomy table
tax_tab_peakpoc <- as.matrix(read.table("ASVs_PeakPOC_Taxa.tsv", header=T, row.names=1, check.names=F, sep = "\t"))

#Load sample matadata file
sample_info_tab_peakpoc <- read.csv("metadata_EUK_new.csv", header = T, row.names = NULL, check.names = F, sep = ",")
row.names(sample_info_tab_peakpoc) <- sample_info_tab_peakpoc$sample_name # Assign rownames from one of the column "sample_name"
sample_info_tab_peakpoc <- sample_info_tab_peakpoc[-4,] # Remove the column since the column is assigned to the rownames. 

#Check content of the metadata file
sample_info_tab_peakpoc

```

- **sep** = a string used to separate columns. using sep = "\t" gives tab deliminated output
- **quote** = a logical value or a numeric vector. If TRUE, any character or factor coluns will be surrounded by double quotes. If FALSE, nothing is quoted/
- **col.names** = either a logical value indicating whether the column names of x are to be written along with x, or a character vector of column names to be written. 
- **row.names** = a vector of row names. Setting it to 1 means the name of samples/ASV number is listed in the first column of the table
- **header** = TRUE if a table contains header for columns. 

## Assigning contigency tables and sample metadata into a phyloseq object

```{r}

#read ASV table into OTU_table class object
otu_1_peakpoc <- as.matrix(count_tab_peakpoc) #Load ASV table
otu_2_peakpoc <- decostand(count_tab_peakpoc, MARGIN = 2, method = "hellinger" ) # transform a count table with hellinger transformation
otu_i_peakpoc <- otu_table(otu_1_peakpoc, taxa_are_rows = T) # Count table
otu_d_peakpoc <- otu_table(otu_2_peakpoc, taxa_are_rows = T) # Count table transformed by hellinger transformation
rm(otu_1_peakpoc)

#read taxonomy table into tax_table class object
TAX_peakpoc <- tax_table(tax_tab_peakpoc)

#read metadata file into sample_data class object
samples_peakpoc <- sample_data(sample_info_tab_peakpoc)

#Construct phyloseq object from the component objects above. 
Ps_peakpoc <- phyloseq(otu_i_peakpoc, TAX_peakpoc, samples_peakpoc)
Ps_hellinger <- phyloseq(otu_d_peakpoc, TAX_peakpoc, samples_peakpoc)

#Check content of the object
Ps_peakpoc
Ps_hellinger

#Change column names to sample names

sample_data(Ps_peakpoc)$sample_ID <- factor(sample_data(Ps_peakpoc)$sample_ID, 
                                            levels = sample_data(Ps_peakpoc)$sample_ID[order(as.factor(as.Date(sample_data(Ps_peakpoc)$collection_date_end, format = "%m/%d/%Y")))])

Ps_peakpoc_merged <- merge_samples(Ps_peakpoc, group = "sample_thesis") # This is only neccesary if you are not using plotting functions by phyloseq (e.g. dendrogram), otherwise, the name of samples can be manipulated through a parameter in phyloseq plotting function

```

-- **merge_samples()** = merge/agglomerate the sample indices of a phyloseq object according to a categorical variable contained in a sample_data or a provided factor. group = a single character string mathcin a variable name in the correspoding sample date of x.

## Plotting Chao1 richness estimator, Shannon and Simpson diversity estimator. 

Chao1 estimator is measured by the number of observed species plus number of singletons divided by two times of number of doubletons. Chao1 estimates the number of species that are represented only by a single indivudal or by two inviduals in that sample based on abundances.

Shannon diversity estimator shows both richness and evenness of species abundance. In other words, if a sample has high shannon diversity index, the sample is thought to be more 'diverse' compared to those samples with lower diversity index 

Simpson's diversity index is a measure of diversity which takes into account the number of species present, as well as the relative abundance of each species. 

```{r}
#plot alpha diversity by samples

plot_richness(Ps_peakpoc, measures = c("Chao1", "Shannon", "Simpson"), x = "sample_thesis", color = "season") + 
  geom_point(size = 5) + # Enlarge data points 
  theme(axis.text.x = element_text(size = 9, angle = 90, face = "bold")) + # Increase the size of axis text 
  labs(x = "Sample Name") +
  theme(legend.title = element_text(size = 14),
        legend.text = element_text(size = 12),
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 0.6)
  )

```


```{r}
# plot alpha diversity by month-

# This is for ggplot to recognize and chronologically sort dates in "collection_date_end", 
sample_data(Ps_peakpoc)$collection_date_end <- as.factor(as.Date(sample_data(Ps_peakpoc)$collection_date_end, format = "%m/%d/%Y"))

# Plot based on the collection date. 
plot_richness(Ps_peakpoc, measures = c("Chao1", "Shannon", "Simpson"), x="collection_date_end", color="season") + 
  labs(x = "Collection date end") + 
  geom_point(size = 5) +
  theme(axis.text.x = element_text(size = 10, angle = 90, face = "bold")) + # Increase the size of axis text 
  labs(x = "Collection Date") +
  theme(legend.title = element_text(size = 14),
        legend.text = element_text(size = 12),
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 0.6)
  )


```

## iNEXT rarefaction curve

iNEXT function only allws data frame or contigency table as input, so if samples were grouped or names of columns were changed in phyloseq, it is important to transform a phyloseq oject into a data frame. If you wish to change the column names of the original table to the specific name which taken from sample data table in phyloseq object, use a phyloseq object created in **line 126**

```{r}
# Exporting a count table with desired sample names from a phyloseq object 
otu_table_site <- as(otu_table(Ps_peakpoc_merged), "matrix") 
if(!taxa_are_rows(Ps_peakpoc_merged)){otu_table_site <- t(otu_table_site)}
otu_table_site_df_peakpoc <- as.data.frame(otu_table_site)

# Check the content
otu_table_site_df_peakpoc

```

```{r}

# rarefaction curves using iNEXT
out <- iNEXT(otu_table_site_df_peakpoc, q=0, datatype = "abundance")
levels(out$DataInfo$site) = sample_data(Ps_peakpoc)$sample_ID[order(as.Date(sample_data(Ps_peakpoc)$collection_date_end, format = "%m/%d/%Y"))]

# Plot 
ggiNEXT(out, type=1, se= TRUE, facet.var = "order", color = "site") + theme(legend.text = element_text(size = 7)) + ylab("Number of ASVs") + 
  theme(legend.title = element_text(size = 14),
        legend.text = element_text(size = 12, face = "bold"),
        axis.text.x = element_text(size = 10, face = "bold"),
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 0.6)
  ) + 
  scale_shape_manual(values=c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,0,1,2,3,4))



ggiNEXT(out, type=2, se = TRUE, facet.var = "order", color.var = "site") + theme(legend.text = element_text(size = 7)) + ggtitle("Sample completeness curve")
ggiNEXT(out, type=3, se = TRUE, facet.var = "order", color.var = "site") + theme(legend.text = element_text(size = 7)) + ggtitle("Coverage based curve")
```

There are three types of curves plotted above.

- **Type 1** = Sample- size based R/E curve. This curve plots diversity estimates with confidence intervals as a function of sample size up to double the reference sample size.
- **Type 2** = sample completeness curve with confidence intervals. This curve plots the sample coverage with respect to sample size for the same range with type 1 plot
- **Type 3** = Coverage based R/E curve, This curve plots the diversity estimates with confidence intervals as a function of sample coverage up to the maximum coverage obrained from the maximum size described in type 1
- **knots** = an integer specifying the numver of equally-spaced knots between size 1 andthe endpoint. Since my data set is large and the compational time takes forever, the knots parameter is set to 100 to draw the rarefaction curves. 
- **conf** = a positive number <1 specifying the level of condence interval. I do not wish to calculate the confidence interval. Therefore, the value is set to NULL. 
- **se** = if TRUE, it plots confidence intervals. 
- **theme()** = Changing legend text font size. 
- **ggtitle()** = add title to plot. 

## Community Compositions

Ratio transformations capture the relationships between each ASVs in the dataset and these ratios are the same whether the data are counts or proportions. Taking the logarithm of these ratios makes the data symmetric and linearly related and places the data in a log ratio coordinate space. 

```{r}
# export a count table with the right column names
count_peakpoc_correct_names <- as(otu_table(Ps_peakpoc_merged), "matrix") # extract a count table from the phyloseq object
merge_peakpoc <- merge(t(count_peakpoc_correct_names), tax_tab_peakpoc, by = 0, sort = FALSE) # combine the count table and taxa table to have rownames with taxa names.  

# Aggregate ASV counts by Phylum
clr_phylum_peakpoc <-  merge_peakpoc %>% 
  dplyr::select(2:25, 28) %>% # select() extracts columns from input number.  
  group_by(Phylum) %>% # aggreagate counts by phylum
  summarise_all(funs(sum)) 

# Removing Incertae Sedis	
clr_phylum_peakpoc <- clr_phylum_peakpoc[-7,] 

# Checking sum of the counts before and after the aggregation. 

colSums(clr_phylum_peakpoc[,-1]) # sum of aggregated table
colSums(t(count_peakpoc_correct_names[,-1])) # sum of original table


# Make taxonomy name as rownames
row.names(clr_phylum_peakpoc) <- clr_phylum_peakpoc$Phylum
clr_phylum_peakpoc$Phylum <- NULL

# make samples into rows
clr_phylum_peakpoc_t <- t(clr_phylum_peakpoc)

# if the value in the table is 0, add a small constant 0.1 to make sure we can take the log
clr_phylum_peakpoc_t[clr_phylum_peakpoc_t == 0] <- 0.1

# CLR transformation on OTU table
# CLR = taking log from ratio calculated by proportion of each ASVs within a sample

rm(clr_phylum_peakpoc)
clr_phylum_peakpoc <- t(clr(clr_phylum_peakpoc_t, ifwarn = FALSE))
rm(clr_phylum_peakpoc_t)

# Repeat same steps for Class, Order, and Family. In this block, I will use source() to reduce redundancy

source("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\PeakPOC\\clr_class_order_family.R")
```
- **by** = specifications of the columns used for merging. By default the data frames are merged on the columns with names they both have. In this case, rownames from the two tables need to match, so **0** is used in this parameter.
- **sort** = logical. Should the result be sorted on the by columns? 
- **ifwarn** = by default ifwarn = TRUE which generates a reminder/warning that when carrying out a centred log-ratio transformation all the data/parts must be in the same measurement units. The message can be suppressed by setting ifwarn = FALSE.

**Cross checking**
When plotting comunity compositions by each taxa rank (phylum, class, order, and family), the counts that corespond to identical taxonomy must be aggregated in order to avoid redundancy in the dataset. The original data table is filtered into each taxa rank with aggregated ASV counts and the data is stored in a object. Correct taxonomy names and its number of counts are manually cross checked by looking at first few rows in the object and the original data table.   


Now we can plot. 
```{r}
# plot

source("C:\\Users\\MSI\\AppData\\Local\\Packages\\CanonicalGroupLimited.UbuntuonWindows_79rhkp1fndgsc\\LocalState\\rootfs\\home\\matomo\\jacobs\\BCCB_Thesis\\PeakPOC\\plot_community_compositions.R")

```

## Hiracheal clustering dendrogram
```{r}
# Create a phyloseq object with values from a variable "sample_ID" as sample names. 
Ps_sample_name_peakpoc <- merge_samples(Ps_hellinger, "sample_thesis")

#Caluclate distances
d_peakpoc <- phyloseq::distance(Ps_sample_name_peakpoc, method = "bray", type = "samples")

# Hierarchical cluster analysis on a set of dissimilarities
p.hclust_peakpoc <- as.dendrogram(hclust(d_peakpoc, method = "average"))

# color scheme
color_peakpoc <- c("red", "red", "red", "red", "red", "red", "red", "red", "blue", "blue", "red", "blue", "red", "blue", "blue", "red", "blue", "red", "blue", "blue", "red", "blue", "blue", "blue")
labels_colors(p.hclust_peakpoc) <- color_peakpoc
labels_cex(p.hclust_peakpoc) = 0.75
plot(p.hclust_peakpoc, hang = -3 ,cex.lab = 1.2, cex.axis = 1, cex = 1, ylim=c(0.0, 1.0), ylab = "Bray-Curtis Dissimilarity")
```


## NMDS ordination plot

NMDS is a way to condese information from mulidementional data into a 2D representation or ordination. The closer two points are, the more similar the corresponding samples are with respect to the variables that went into making the NMDS plot. 

```{r}
# Transpose a count tab (because vegan accepts rows as samples and columns as species.)
t_count_peakpoc <- t(count_tab_peakpoc)

# Hellinger transformation
NMDS_peakpoc <- decostand(t_count_peakpoc, MARGIN = 2, method = "hellinger" )

# Ordinate
peakpoc_NMDS <- vegan::metaMDS(NMDS_peakpoc, k=3, distance = "bray")

peakpoc_NMDS$stress

# Plot stress plot. 
stressplot(peakpoc_NMDS)
```

```{r}
# By season

spring <- subset_samples(Ps_peakpoc, season == "spring")
summer <- subset_samples(Ps_peakpoc, season == "summer")
count_peakpoc_spring <- as(otu_table(spring), "matrix")
count_peakpoc_summer <- as(otu_table(summer), "matrix")

# Transpose a count tab (because vegan accepts rows as samples and columns as species.)
t_count_peakpoc_spring <- t(count_peakpoc_spring)
t_count_peakpoc_summer <- t(count_peakpoc_summer)

# Hellinger transformation
NMDS_spring <- decostand(t_count_peakpoc_spring, MARGIN = 2, method = "hellinger")
NMDS_summer <- decostand(t_count_peakpoc_summer, MARGIN = 2, method = "hellinger")

# Ordinate
spring_NMDS <- vegan::metaMDS(NMDS_spring, k=3, distance = "bray")
summer_NMDS <- vegan::metaMDS(NMDS_summer, k=2, distance = "bray")

spring_NMDS$stress
summer_NMDS$stress

# Plot stress plot. 
stressplot(spring_NMDS)
stressplot(summer_NMDS)
```

- **MARGIN** = Margin, if default is not accept. "1" = rows, and "2" = columns of x
- **k** = Number of dimensions. The number od points n should be n>2*k+1 and prefeably higher in non-metric MDS
- **distance** = Dissimilarity index used in vegdist. We will use bray curtis dissimilarity index. 

OKay now moving on to plotting the points

```{r}

# Overview of our data
plot(peakpoc_NMDS)

# extract NMDS scores (x and y coordinates)
data.scores_peakpoc <- as.data.frame(scores(peakpoc_NMDS))

# add columns to data frame
data.scores_peakpoc$collection_date_end = as.Date(strptime(sample_info_tab_peakpoc$collection_date_end, '%m/%d/%Y'))
data.scores_peakpoc$year = format(data.scores_peakpoc$collection_date_end, "%Y") # This is to add points by year
data.scores_peakpoc$season = sample_info_tab_peakpoc$season
data.scores_peakpoc$month = sample_info_tab_peakpoc$month
data.scores_peakpoc$Chla_mean_.satellite_HG_region. = sample_info_tab_peakpoc$Chla_mean_.satellite_HG_region.
  
# Check the content of the metadata
head(data.scores_peakpoc)


# plotting the NMDS.
NMDS_new_peakpoc <- ggplot(data.scores_peakpoc, aes(x = NMDS1, y= NMDS2)) +
  geom_point(size = 4, aes( shape = season, color = year)) + # Choosing mapping for the plot
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
  axis.text.x = element_text(colour = "black", face = "bold", size = 12), # axis text formatting
  legend.text = element_text(size = 12, face ="bold", colour ="black"), # legend text formatting
  legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), # legend position formatting
  axis.title.x = element_text(face = "bold", size = 14, colour = "black"), # axis title formatting
  legend.title = element_text(size = 12, colour = "black", face = "bold"), # legend title formatting
  panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), # change background color
  legend.key=element_blank()) +
  labs(color = "Collection year", shape = "Season") #  Label formatting
  
# color code
# cc <- scales::seq_gradient_pal("lightblue2", "midnightblue")(seq(0,1,length.out=12))
  NMDS_new_peakpoc 
  # + scale_color_manual(values=cc)
```

## Analyze by season

```{r}
# extract NMDS scores (x and y coordinates)
data.scores_spring <- as.data.frame(scores(spring_NMDS))
data.scores_summer <- as.data.frame(scores(summer_NMDS))

# add columns to data frame
data.scores_spring$collection_date_end = as.Date(strptime(sample_data(spring)$collection_date_end, '%Y-%m-%d'))
data.scores_spring$year = format(data.scores_spring$collection_date_end, "%Y") # This is to add points by year
data.scores_summer$collection_date_end = as.Date(strptime(sample_data(summer)$collection_date_end, '%Y-%m-%d'))
data.scores_summer$year = format(data.scores_summer$collection_date_end, "%Y") # This is to add points by year

# Check the content of the metadata
head(data.scores_spring)
head(data.scores_summer)

# plotting the NMDS.
NMDS_new_spring <- ggplot(data.scores_spring, aes(x = NMDS1, y= NMDS2)) +
  geom_point(size = 4, aes(color = year)) + # Choosing mapping for the plot
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
  axis.text.x = element_text(colour = "black", face = "bold", size = 12), # axis text formatting
  legend.text = element_text(size = 12, face ="bold", colour ="black"), # legend text formatting
  legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), # legend position formatting
  axis.title.x = element_text(face = "bold", size = 14, colour = "black"), # axis title formatting
  legend.title = element_text(size = 12, colour = "black", face = "bold"), # legend title formatting
  panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), # change background color
  legend.key=element_blank()) +
  labs(color = "Collection year") #  Label formatting
  
NMDS_new_summer <- ggplot(data.scores_summer, aes(x = NMDS1, y= NMDS2)) +
  geom_point(size = 4, aes(color = year)) + # Choosing mapping for the plot
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
  axis.text.x = element_text(colour = "black", face = "bold", size = 12), # axis text formatting
  legend.text = element_text(size = 12, face ="bold", colour ="black"), # legend text formatting
  legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), # legend position formatting
  axis.title.x = element_text(face = "bold", size = 14, colour = "black"), # axis title formatting
  legend.title = element_text(size = 12, colour = "black", face = "bold"), # legend title formatting
  panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), # change background color
  legend.key=element_blank()) +
  labs(color = "Collection year") #  Label formatting
  
# color code
 cc <- scales::seq_gradient_pal("pink", "red")(seq(0,1,length.out=12))
  NMDS_new_spring  + scale_color_manual(values=cc)
  
  cc2 <- scales::seq_gradient_pal("lightblue", "darkblue")(seq(0,1,length.out=12))
  NMDS_new_summer + scale_color_manual(values=cc2)

```






## Redundancy analysis 

Redundancy analysis (RDA) is a method to extract and summarise the variation in a set of response variables that can be explained by a set of explanatory 
variables. More accurately, RDA is a direct gradient analysis technique which summarises linear relationships between components of response variables that are 
"redundant" with (i.e. "explained" by) a set of explanatory variables.

```{r}

# Reorder sample name by date

sample_data(Ps_hellinger)$sample_ID <- factor(sample_data(Ps_hellinger)$sample_ID, 
                                            levels = sample_data(Ps_hellinger)$sample_ID[order(as.factor(as.Date(sample_data(Ps_hellinger)$collection_date_end, format = "%m/%d/%Y")))])

sample_data(Ps_hellinger)$collection_date_end <- as.Date(sample_data(Ps_hellinger)$collection_date_end, format = "%m/%d/%Y")
sample_data(Ps_hellinger)$year <- as.factor(format(data.scores_peakpoc$collection_date_end, "%Y")) # This is to add points by year

# ordinate
rda_peakpoc <- ordinate(Ps_hellinger, method = "RDA", distance = "bray", formula = ~ Ice + Chlorophyll_a + Temperature)

# Plot

RDA_peakpoc <- plot_ordination(Ps_hellinger, rda_peakpoc, type = "samples", color="year", shape = "season")

# Check the plot
RDA_peakpoc


# Now add the environmental variables as arrows
arrowmat_peakpoc = vegan::scores(rda_peakpoc, display = "bp")

# Add labels, make a data.frame
arrowdf_peakpoc <- data.frame(labels = rownames(arrowmat_peakpoc), arrowmat_peakpoc)

# Define the arrow aesthetic mapping
arrow_map = aes(xend = RDA1, yend = RDA2, x = 0, y = 0, shape = NULL, color = NULL, label = labels)
arrowhead = arrow(length = unit(0.05, "npc")) # length of arrows
label_map <- aes(x = 0.8 * RDA1,
    y = 1.2 * RDA2, 
    shape = NULL, 
    color = NULL, 
    label = labels)

cc_peakpoc <- scales::seq_gradient_pal("green", "purple")(seq(0,1,length.out=12))

# Make a new graphic
RDA_peakpoc + 
  geom_point(size = 5) + 
  geom_segment(
    mapping = arrow_map, 
    size = 1, 
    data = arrowdf_peakpoc, 
    color = "darkgrey", 
    arrow = arrowhead
  ) + 
  geom_text(
    mapping = label_map, 
    size = 5,  
    data = arrowdf_peakpoc, 
    show.legend = FALSE
  ) +
   theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), # axis text formatting
    legend.text = element_text(size = 12, face ="bold", colour ="black"), # legend text formatting
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), # legend position formatting
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), # axis title formatting
    legend.title = element_text(size = 12, colour = "black", face = "bold"), # legend title formatting
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), # change background color
    legend.key=element_blank()) + 
    labs( color = "Year",  shape = "Season") + # Label formatting
    scale_color_manual(values=cc_peakpoc)
```


```{r}
# by season
spring_RDA <- subset_samples(Ps_hellinger, season == "spring")
summer_RDA <- subset_samples(Ps_hellinger, season == "summer")

# Reorder sample name by date

sample_data(spring_RDA)$collection_date_end <- as.Date(sample_data(spring_RDA)$collection_date_end, format = "%Y-%m-%d")
sample_data(spring_RDA)$year <- as.factor(format(sample_data(spring_RDA)$collection_date_end, "%Y")) # This is to add points by year

sample_data(summer_RDA)$collection_date_end <- as.Date(sample_data(summer_RDA)$collection_date_end, format = "%Y-%m-%d")
sample_data(summer_RDA)$year <- as.factor(format(sample_data(summer_RDA)$collection_date_end, "%Y")) # This is to add points by year

# ordinate
rda_spring <- ordinate(spring_RDA, method = "RDA", distance = "bray", formula = ~ Ice + Chlorophyll_a + Temperature)
rda_summer <- ordinate(summer_RDA, method = "RDA", distance = "bray", formula = ~ Ice + Chlorophyll_a + Temperature)

# Plot

RDA_spring <- plot_ordination(spring_RDA, rda_spring, type = "samples", color="year")
RDA_summer <- plot_ordination(summer_RDA, rda_summer, type = "samples", color="year")

# Now add the environmental variables as arrows
arrowmat_spring = vegan::scores(rda_spring, display = "bp")
arrowmat_summer = vegan::scores(rda_summer, display = "bp")

# Add labels, make a data.frame
arrowdf_spring <- data.frame(labels = rownames(arrowmat_spring), arrowmat_spring)
arrowdf_summer <- data.frame(labels = rownames(arrowmat_summer), arrowmat_summer)

# Define the arrow aesthetic mapping
arrow_map = aes(xend = RDA1, yend = RDA2, x = 0, y = 0, shape = NULL, color = NULL, label = labels)
arrowhead = arrow(length = unit(0.05, "npc")) # length of arrows
label_map <- aes(x = 0.8 * RDA1,
    y = 1.2 * RDA2, 
    shape = NULL, 
    color = NULL, 
    label = labels)

# Make a new graphic
RDA_spring + 
  geom_point(size = 5) + 
  geom_segment(
    mapping = arrow_map, 
    size = 1, 
    data = arrowdf_spring, 
    color = "darkgrey", 
    arrow = arrowhead
  ) + 
  geom_text(
    mapping = label_map, 
    size = 5,  
    data = arrowdf_spring, 
    show.legend = FALSE
  ) +
   theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), # axis text formatting
    legend.text = element_text(size = 12, face ="bold", colour ="black"), # legend text formatting
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), # legend position formatting
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), # axis title formatting
    legend.title = element_text(size = 12, colour = "black", face = "bold"), # legend title formatting
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), # change background color
    legend.key=element_blank()) + 
    labs( color = "Year") + # Label formatting
    scale_color_manual(values=cc)

# Make a new graphic
RDA_summer + 
  geom_point(size = 5) + 
  geom_segment(
    mapping = arrow_map, 
    size = 1, 
    data = arrowdf_summer, 
    color = "darkgrey", 
    arrow = arrowhead
  ) + 
  geom_text(
    mapping = label_map, 
    size = 5,  
    data = arrowdf_summer, 
    show.legend = FALSE
  ) +
   theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), # axis text formatting
    legend.text = element_text(size = 12, face ="bold", colour ="black"), # legend text formatting
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), # legend position formatting
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), # axis title formatting
    legend.title = element_text(size = 12, colour = "black", face = "bold"), # legend title formatting
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), # change background color
    legend.key=element_blank()) + 
    labs( color = "Year") + # Label formatting
    scale_color_manual(values=cc2)
```


## **MDS ordination plot**

```{r}

#Calculate distance
Ps_MDS_peakpoc <- ordinate(Ps_hellinger, "MDS", "bray")

#plot

#by sample sites
MDS_peakpoc <- plot_ordination(Ps_hellinger, Ps_MDS_peakpoc, color = "year", shape = "season") +
  geom_point(size = 4) + # Choosing mapping for the plot 
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), # Text formatting
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), # axis text formatting
    legend.text = element_text(size = 12, face ="bold", colour ="black"), # legend text formatting
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), # legend position formatting
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), # axis title formatting
    legend.title = element_text(size = 12, colour = "black", face = "bold"), # legend title formatting
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2), # change background color
    legend.key=element_blank()) + 
    labs(color = "Year", shape = "Season") + # Label formatting
    scale_color_manual(values=cc)
MDS_peakpoc

```